---
title: "Data types, structures and loops"
subtitle: "Data Science Flipped Classroom - Module 6"
author: "Adriana Clavijo Daza"
date: "2026-06-04"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    chalkboard: true
footer: "DSFC - Module 6"
editor: source
execute:
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

# Quarto refresher {background-color="#4C326A"}

## The three components of a `.qmd` file

::: {.incremental}
- **YAML header**: metadata and options between `---` fences at the top
- **Markdown text**: headings, prose, lists, links
- **Code chunks**: executable R code between ```` ```{r} ```` fences
:::

. . .

Quarto runs the chunks, weaves the results into the text, and renders everything to HTML, PDF, or slides.

## The YAML header

```yaml
---
title: "My analysis"
author: "Adriana Clavijo Daza"
date: "2026-06-04"
format: html
execute:
  echo: true
---
```

. . .

::: {.task}
Indentation matters in YAML. Two spaces, never tabs.
:::

## Code chunk options

Options go inside the chunk, each on its own line starting with `#|`:

````
```{{r}}
#| label: my-chunk
#| echo: true
#| warning: false

mean(c(1, 2, 3))
```
````

## The most useful chunk options

| Option | Effect |
|--------|--------|
| `label` | names the chunk, useful for navigation and errors |
| `echo` | show the code itself |
| `eval` | run the code |
| `include` | show any output at all |
| `warning` | show warnings |
| `message` | show messages |
| `error` | render even if the chunk errors |

## Options for cross-references

Remember from module 5: two options turn a chunk's output into a [numbered, linkable]{.highlight-peach} table or figure.

````
```{{r}}
#| label: fig-height-mass
#| fig-cap: "Height vs mass of Star Wars characters"

starwars |>
  ggplot(aes(height, mass)) +
  geom_point()
```
````

. . .

::: {.incremental}
- `label` **must** start with `fig-` (figures) or `tbl-` (tables)
- Pair it with `fig-cap` or `tbl-cap`
- Refer to it in the text with `@fig-height-mass`
:::

# Data types in R {background-color="#4C326A"}

## The four basic types

Every value in R has a [type]{.highlight-peach}. The four you will meet daily:

| Type | Examples |
|------|----------|
| `character` | `"hello"`, `"42"` |
| `integer` | `1L`, `42L` |
| `double` | `1.5`, `42`, `pi` |
| `logical` | `TRUE`, `FALSE`, `NA` |

. . .

::: {.task}
`integer` and `double` together are the **numeric** types.
:::

## Checking the type with `typeof()`

```{r}
#| label: typeof-basics
#| echo: true
#| eval: true

typeof("hello")
typeof(1L)
typeof(1.5)
typeof(TRUE)
```

## Integer or double?

A plain number is a [double]{.highlight-peach} by default, even without decimals:

```{r}
#| label: double-default
#| echo: true
#| eval: true

typeof(42)
```

. . .

The `L` suffix asks for an integer explicitly:

```{r}
#| label: integer-suffix
#| echo: true
#| eval: true

typeof(42L)
typeof(1:10)
```

. . .

::: {.task}
Sequences made with `:` are integers. In practice the difference rarely matters; R converts between them silently.
:::

## Coercion: mixing types

A vector can hold [only one type]{.highlight-peach}. When you mix, R converts everything to the most flexible type:

```{r}
#| label: coercion
#| echo: true
#| eval: true

c(1, "two", TRUE)
c(1, TRUE, FALSE)
```

. . .

The hierarchy: `logical` → `integer` → `double` → `character`.

. . .

::: {.task}
This is why a single text entry in a spreadsheet column can turn a whole column of numbers into text after import.
:::

## Explicit conversion

The `as.*()` family converts on demand:

```{r}
#| label: as-family
#| echo: true
#| eval: true

as.numeric("3.14")
as.character(42)
as.integer(TRUE)
```

. . .

Impossible conversions give `NA` with a warning:

```{r}
#| label: as-na
#| echo: true
#| eval: true
#| warning: true

as.numeric("hello")
```

# Data classes {background-color="#4C326A"}

## Type vs class

`typeof()` tells you how a value is [stored]{.highlight-peach}.

`class()` tells you how a value [behaves]{.highlight-peach}.

```{r}
#| label: type-vs-class
#| echo: true
#| eval: true

d <- as_date("2026-06-04")
typeof(d)
class(d)
```

. . .

A `Date` is stored as a double (days since 1970), but its class makes it print and compute like a calendar date.

## The factor class

A [factor]{.highlight-peach} represents a categorical variable: a value that can only take one of a fixed set of [levels]{.highlight-peach}.

```{r}
#| label: factor-basic
#| echo: true
#| eval: true

sizes <- factor(c("small", "large", "small", "medium"))
sizes
levels(sizes)
```

## How factors are stored

Under the hood, a factor is an [integer vector]{.highlight-peach} plus a table of levels:

```{r}
#| label: factor-internal
#| echo: true
#| eval: true

typeof(sizes)
class(sizes)
as.integer(sizes)
```

## Why bother with factors?

::: {.incremental}
- Plots and tables follow [your]{.highlight-peach} order, not the alphabet
- Levels with zero observations still show up in counts
- Typos become visible: a fifth level you did not expect
- Models need them to treat a variable as categorical
:::

# Vectors vs lists {background-color="#4C326A"}

## Atomic vectors

The vectors we have used so far are [atomic]{.highlight-peach}: every element has the same type.

```{r}
#| label: atomic-vector
#| echo: true
#| eval: true

heights <- c(172, 167, 96, 202)
heights
```

. . .

That uniformity is what makes vectorized operations possible:

```{r}
#| label: vectorized
#| echo: true
#| eval: true

heights / 100
```

## Lists

A [list]{.highlight-peach} can hold anything: different types, different lengths, even other lists.

```{r}
#| label: list-basic
#| echo: true
#| eval: true

person <- list(
  name    = "Luke",
  height  = 172,
  films   = c("A New Hope", "Return of the Jedi")
)
str(person)
```

## Vector or list?

:::: {.columns}

::: {.column width="50%"}
**Atomic vector**

- One type for all elements
- Elements are single values
- Built with `c()`
- Fast, vectorized math
:::

::: {.column width="50%"}
**List**

- Each element can differ
- Elements can be whole vectors
- Built with `list()`
- The flexible container
:::

::::

. . .

::: {.task}
`str()` is your friend: it shows the structure of any object at a glance.
:::

## Why lists matter

You meet lists constantly, sometimes without noticing:

A data frame **is** a list of equal-length vectors

# Building data frames {background-color="#4C326A"}

## From vectors to a data frame

A data frame is a [list of vectors]{.highlight-peach} of equal length, displayed as columns.

. . .

Two ways to build one from vectors:

::: {.incremental}
- `data.frame()` from base R
- `tibble()` from the tidyverse
:::

## With base R: `data.frame()`

```{r}
#| label: base-dataframe
#| echo: true
#| eval: true

df <- data.frame(
  name   = c("Luke", "Leia", "R2-D2"),
  height = c(172, 150, 96),
  droid  = c(FALSE, FALSE, TRUE)
)
df
```

## With the tidyverse: `tibble()`

```{r}
#| label: tibble-create
#| echo: true
#| eval: true

tb <- tibble(
  name   = c("Luke", "Leia", "R2-D2"),
  height = c(172, 150, 96),
  droid  = c(FALSE, FALSE, TRUE)
)
tb
```

. . .

A tibble **is** a data frame, with a few surprising behaviours removed.

## `data.frame()` vs `tibble()` {.smaller}

| Behaviour | `data.frame()` | `tibble()` |
|-----------|----------------|------------|
| Printing | everything | first 10 rows, types shown |
| Recycling | any length that divides evenly | only length 1 |
| Column names | "fixes" them, `my var` → `my.var` | keeps them as written |
| Use a column just defined | no | yes |

# Accessing columns {background-color="#4C326A"}

## Four ways to reach a column

```{r}
#| label: access-setup
#| echo: true
#| eval: true

tb <- tibble(
  name   = c("Luke", "Leia", "R2-D2"),
  height = c(172, 150, 96)
)
```

. . .

| Syntax | Returns |
|--------|---------|
| `tb$height` | a vector |
| `tb[["height"]]` | a vector |
| `tb["height"]` | a tibble with one column |
| `pull(tb, height)` | a vector, pipe-friendly |

## `$` and `[[`: get the vector

```{r}
#| label: dollar-bracket2
#| echo: true
#| eval: true

tb$height
tb[["height"]]
```

. . .

`[[` also works with a position or a variable holding the name:

```{r}
#| label: bracket2-position
#| echo: true
#| eval: true

tb[[2]]
col <- "height"
tb[[col]]
```

## `[`: keep the data frame

Single brackets [subset]{.highlight-peach} the data frame, so you get a data frame back:

```{r}
#| label: single-bracket
#| echo: true
#| eval: true

tb["height"]
```

. . .

::: {.task}
A frequent source of confusion: `tb["height"]` looks like the column but will not work where a plain vector is expected, for example in `mean()`.
:::

## `pull()`: the pipe-friendly extractor

`pull()` does what `$` does, but fits at the [end of a pipeline]{.highlight-peach}:

```{r}
#| label: pull-pipeline
#| echo: true
#| eval: true

starwars |>
  filter(species == "Droid") |>
  pull(height)
```

. . .

Compare with `select()`, which keeps the data frame:

```{r}
#| label: select-compare
#| echo: true
#| eval: true

starwars |>
  filter(species == "Droid") |>
  select(height)
```

# For loops {background-color="#4C326A"}

## When you need repetition

Much of R is vectorized, so many loops written in other languages are one line here.

. . .

But sometimes you genuinely need to [repeat a computation]{.highlight-peach}: once per file, once per column, once per simulation.

. . .

That is what a `for` loop does.

## The anatomy of a for loop

```{r}
#| label: loop-anatomy
#| echo: true
#| eval: false

for (item in vector) {
  # body: runs once per element,
  # with `item` set to the current element
}
```

. . .

::: {.incremental}
- `item`: a name you choose for the current element
- `vector`: the values to iterate over
- body: the code to repeat, inside `{ }`
:::

## A first loop

```{r}
#| label: loop-first
#| echo: true
#| eval: true

names <- c("Luke", "Leia", "R2-D2")

for (n in names) {
  print(n)
}
```

. . .

::: {.task}
Inside a loop, results are not printed automatically. Use `print()` to see them.
:::

## Looping over positions

Often you want the [index]{.highlight-peach}, not the value. Use `seq_along()`:

```{r}
#| label: loop-seq-along
#| echo: true
#| eval: true

heights <- c(172, 150, 96)

for (i in seq_along(heights)) {
  print(heights[i] / 100)
}
```

. . .

::: {.task}
Prefer `seq_along(x)` over `1:length(x)`. If `x` is empty, `1:0` gives `c(1, 0)` and the loop runs twice instead of not at all.
:::

## Collecting results

Create an [output container first]{.highlight-peach}, then fill it position by position:

```{r}
#| label: loop-collect
#| echo: true
#| eval: true

heights <- c(172, 150, 96)
heights_m <- vector()

for (i in seq_along(heights)) {
  heights_m[i] <- heights[i] / 100
}

heights_m
```

## Looping over columns

A data frame is a list, so you can loop over its columns:

```{r}
#| label: loop-columns
#| echo: true
#| eval: true

tb <- tibble(height = c(172, 150, 96), mass = c(77, 49, 32))
means <- vector("double", ncol(tb))

for (i in seq_along(tb)) {
  means[i] <- mean(tb[[i]])
}

means
```

. . .

Note the `[[`: each column comes out as a plain vector.

## Loop or vectorize?

::: {.incremental}
- If a vectorized function exists, use it: `heights / 100` beats any loop
- Loops shine when steps depend on each other, or for side effects like saving one file per group
:::

## Learn more

::: {.learn-more}
[https://quarto.org/docs/computations/execution-options.html](https://quarto.org/docs/computations/execution-options.html)
:::

::: {.learn-more}
[https://r4ds.hadley.nz/base-R](https://r4ds.hadley.nz/base-R)
:::

::: {.learn-more}
[https://r4ds.hadley.nz/factors](https://r4ds.hadley.nz/factors)
:::

::: {.learn-more}
[https://r4ds.hadley.nz/iteration](https://r4ds.hadley.nz/iteration)
:::

## {background-color="#4C326A"}

::: {.hand-purple-large}
Questions?
:::
