---
title: "Data visualization"
subtitle: "Data Science Flipped Classroom - Module 2"
author: "Adriana Clavijo Daza"
date: "2026-04-02"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    chalkboard: true
footer: "DSFC - Module 2"
editor: source
execute:
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggridges)
```

# ggplot2 {background-color="#4C326A"}

## What is ggplot2?

ggplot2 is an R package for [data visualization]{.highlight-peach} based on the **Grammar of Graphics**.

. . .

It lets you build plots by combining components in a structured way, like building sentences.

## Grammar of Graphics

Example:

. . .

> "Make a bar chart of sales by region and products."

. . .

In a structured way, translates to:

> "Using the sales data, put region on the x-axis, sales on the y-axis, and group bars by product with a unique color for each region."

. . .

::: {.hand-purple}
Each part of the sentence maps to a component of the plot.
:::

# Essential building blocks {background-color="#4C326A"}

## The three essentials

::: {.incremental}
1. **Data**: a `data.frame` or `tibble`, passed to `ggplot()`
2. **Aesthetics**: how variables map to visual properties (position, color, size), defined inside `aes()`
3. **Layers**: where data takes shape (points, lines, bars), added with `geom_*()` or `stat_*()` functions
:::

. . .

::: {.task}
**Good to know**

With just these three elements you can already generate a plot!
:::

## The `mpg` dataset

`mpg` is a dataset included in ggplot2 with fuel economy data for 234 cars from 1999 to 2008.

```{r}
#| label: mpg-preview
#| echo: true
#| eval: true

mpg |>
  select(manufacturer, model, displ, year, cyl, hwy) |>
  head(5)
```

## Example: a basic plot

```{r}
#| label: basic-plot
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

## Breaking it down

```r
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

. . .

| Component | Code | Role |
|-----------|------|------|
| Data | `data = mpg` | The dataset |
| Aesthetics | `aes(x = displ, y = hwy)` | Map variables to axes |
| Layer | `geom_point()` | Draw points |

# Additional components {background-color="#4C326A"}

## Scales

[Scales]{.highlight-peach} fine-tune how data is visually represented.

. . .

With `scale_*()` functions you can:

::: {.incremental}
- Change color palettes and shapes
- Modify axis labels and limits
- Control legend behavior
:::

## Coordinates

[Coordinates]{.highlight-yellow} define the spatial context of the plot.

. . .

With `coord_*()` functions you can:

::: {.incremental}
- Apply spatial transformations (e.g., `coord_flip()`)
- Set aspect ratios
- Zoom into a region of the plot
:::

## Facets

[Facets]{.highlight-peach} split a plot into multiple panels based on subsets of the data.

. . .

::: {.incremental}
- `facet_wrap()`: panels from one variable
- `facet_grid()`: panels from two variable combinations
:::

. . .

::: {.hand-purple}
Useful to uncover patterns across groups!
:::

## Themes

[Themes]{.highlight-yellow} control the overall visual style of the plot.

. . .

::: {.incremental}
- `theme_*()` functions apply a complete style (e.g., `theme_minimal()`)
- `theme()` lets you fine-tune individual elements (text, backgrounds, gridlines)
:::

## The full "sentence"

```{r}
#| label: full-plot
#| echo: true
#| eval: true
#| fig-align: center
#| code-line-numbers: "|1|2|3|4|5|6|7"
#| output-location: slide

ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  scale_x_continuous(name = "Engine displacement") +
  scale_y_continuous(name = "Highway MPG") +
  scale_color_brewer(name = "Class", palette = "Set2") +
  facet_wrap(~drv) +
  theme_minimal()
```

## All the components together {.smaller}

| Component | Function | Purpose |
|-----------|----------|---------|
| Data | `ggplot(data)` | The dataset to visualize |
| Aesthetics | `aes()` | Map variables to visual properties |
| Layers | `geom_*()` / `stat_*()` | Geometric shapes or summaries |
| Scales | `scale_*()` | Fine-tune visual mappings |
| Coordinates | `coord_*()` | Spatial transformations |
| Facets | `facet_*()` | Split into panels |
| Themes | `theme_*()` / `theme()` | Visual style |

# Common plot types {background-color="#4C326A"}

## Scatter plot: `geom_point()`

```{r}
#| label: geom-point
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

## Line plot: `geom_line()`

```{r}
#| label: geom-line
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()
```

## Bar chart: `geom_bar()`

```{r}
#| label: geom-bar
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

ggplot(mpg, aes(x = class)) +
  geom_bar()
```

## Column chart: `geom_col()`

```{r}
#| label: geom-col
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

mpg |>
  count(class) |>
  ggplot(aes(x = class, y = n)) +
  geom_col()
```

::: {.task}
`geom_bar()` counts for you, `geom_col()` uses values you provide.
:::

## Histogram: `geom_histogram()`

```{r}
#| label: geom-histogram
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2)
```

## Box plot: `geom_boxplot()`

```{r}
#| label: geom-boxplot
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot()
```

## Smooth line: `geom_smooth()`

```{r}
#| label: geom-smooth
#| echo: true
#| eval: true
#| fig-align: center
#| message: false
#| output-location: slide

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```

::: {.task}
You can layer multiple `geom_*()` functions on the same plot!
:::

## Ridge plot: `geom_density_ridges()`

```{r}
#| label: geom-ridges
#| echo: true
#| eval: true
#| fig-align: center
#| message: false
#| output-location: slide

ggplot(mpg, aes(x = hwy, y = class)) +
  geom_density_ridges()
```

::: {.task}
Ridge plots are great for comparing distributions across groups. Requires the `ggridges` package.
:::

## Summary with `stat_summary()`

```{r}
#| label: stat-summary
#| echo: true
#| eval: true
#| fig-align: center
#| output-location: slide

ggplot(mpg, aes(x = class, y = hwy)) +
  stat_summary(fun = mean, geom = "col") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2)
```

::: {.task}
`stat_*()` functions compute summaries before plotting.
:::

# Choosing the right plot {background-color="#4C326A"}

## What plot should I use?

It depends on the [type of variable]{.highlight-peach} you want to visualize.

## Types of variables

:::: {.columns}

::: {.column width="50%"}
**Non-numerical**

::: {.incremental}
- **Categorical**: no inherent order (e.g., color, country)
- **Ordinal**: with a meaningful order (e.g., education level, rating)
:::
:::

::: {.column width="50%"}
**Numerical**

::: {.incremental}
- **Discrete**: countable values (e.g., number of children)
- **Continuous**: any value in a range (e.g., temperature, weight)
:::
:::

::::

## Can you tell the type?

::: {.hand-purple}
Does it make sense to do math with this variable?
:::

## Zip code: 10115, 28001, 75008...

. . .

Looks like a number, but...

. . .

What is the **average zip code**?

. . .

It makes no sense. Zip codes are [categorical]{.highlight-peach}.

## Year: 2019, 2020, 2021...

. . .

It depends on the context!

. . .

- Comparing groups (sales in 2020 vs 2021): [categorical]{.highlight-peach}
- Tracking change over time: [continuous]{.highlight-yellow}

. . .

::: {.task}
The same variable can be treated differently depending on your question.
:::

## Likert scale: 1, 2, 3, 4, 5

. . .

A satisfaction survey asks: "Rate from 1 (very bad) to 5 (excellent)."

. . .

Is the difference between 1 and 2 the same as between 4 and 5?

. . .

Not necessarily. These are [ordinal]{.highlight-peach}, not truly numerical.

## Age group: "18-25", "26-35", "36-45"...

. . .

The values contain numbers, but...

. . .

You can't average "18-25" and "26-35".

. . .

These are [ordinal]{.highlight-peach}: categories with a meaningful order.

## "How many hours did you sleep?" 4, 6, 7, 8...

. . .

Looks continuous, but people usually answer in whole numbers.

. . .

Is "7 hours" really 7.0, or just a rounded estimate?

. . .

In a survey, this is arguably [discrete]{.highlight-yellow}: the responses are countable, even if the underlying variable is continuous.

## Matching variables to plots {.smaller}

| Variables | Plot type | `geom_*()` |
|-----------|-----------|------------|
| 1 categorical | Bar chart | `geom_bar()` |
| 1 continuous | Histogram | `geom_histogram()` |
| 1 continuous | Density | `geom_density()` |
| 2 continuous | Scatter plot | `geom_point()` |
| 2 continuous | Line plot | `geom_line()` |
| 1 categorical + 1 continuous | Box plot | `geom_boxplot()` |
| 1 categorical + 1 continuous | Ridge plot | `geom_density_ridges()` |
| 1 categorical + 1 continuous | Column chart | `geom_col()` |
| 2 categorical | Stacked/grouped bar | `geom_bar()` |

## {background-color="#4C326A"}

::: {.hand-purple-large}
Questions?
:::
