---
title: "Conditions, dates and tables"
subtitle: "Data Science Flipped Classroom - Module 5"
author: "Adriana Clavijo Daza"
date: "2026-05-21"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    chalkboard: true
footer: "DSFC - Module 5"
editor: source
execute:
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(knitr)
```

# Logical vectors and comparisons {background-color="#4C326A"}

## Logical values

R has a third atomic type alongside numbers and text: [logical]{.highlight-peach}.

. . .

Two possible values, `TRUE` and `FALSE` (plus `NA` for unknown).

```{r}
#| label: logical-basic
#| echo: true
#| eval: true

c(TRUE, FALSE, TRUE)
```

. . .

`T` and `F` work but are not reserved words, so they can be overwritten. Always write the full `TRUE` and `FALSE`.

## Why logical vectors matter

Most data questions reduce to:

. . .

::: {.incremental}
- Which rows match a condition?
- Which values are missing?
- Where is the price above 1000?
:::

. . .

The answer is a logical vector you can pass to `filter()`, `if_else()`, or `case_when()`.

## Comparison operators

| Operator | Meaning |
|----------|---------|
| `<` | less than |
| `<=` | less than or equal to |
| `>` | greater than |
| `>=` | greater than or equal to |
| `==` | equal to |
| `!=` | not equal to |

. . .

::: {.task}
A single `=` is for assignment in function calls. Equality is `==`.
:::

## Comparisons are vectorized

The comparison runs element by element:

```{r}
#| label: cmp-vec
#| echo: true
#| eval: true

x <- c(2, 5, 9, 12)
x > 5
```

. . .

Recycling kicks in if one side is shorter:

```{r}
#| label: cmp-recycle
#| echo: true
#| eval: true

y <- c(1, 6, 8, 10)
x > y
```

## Equality with decimals

`==` on doubles can surprise you:

```{r}
#| label: equal-double
#| echo: true
#| eval: true

0.1 + 0.2 == 0.3
```

. . .

Use `near()` for numerical equality:

```{r}
#| label: equal-near
#| echo: true
#| eval: true

near(0.1 + 0.2, 0.3)
```

## Missing values and `is.na()`

`NA` is contagious. Any comparison with `NA` returns `NA`, not `TRUE` or `FALSE`:

```{r}
#| label: na-cmp
#| echo: true
#| eval: true

NA == NA
NA > 5
```

. . .

To test for missingness, use `is.na()`:

```{r}
#| label: na-test
#| echo: true
#| eval: true

x <- c(1, NA, 3, NA)
is.na(x)
```

## Membership with `%in%`

`%in%` tests whether each value on the left appears anywhere on the right.

```{r}
#| label: in-basic
#| echo: true
#| eval: true

species <- c("Droid", "Human", "Wookiee", "Droid")
species %in% c("Human", "Wookiee")
```

. . .

It is the readable alternative to chaining `==` with `|`:

```{r}
#| label: in-equiv
#| echo: true
#| eval: false

species == "Human" | species == "Wookiee"
```

## Combining conditions

| Operator | Meaning |
|----------|---------|
| `&` | AND, element by element |
| `|` | OR, element by element |
| `!` | NOT |

```{r}
#| label: combine-logical
#| echo: true
#| eval: true

x <- c(2, 5, 9, 12)
x > 3 & x < 10
```

## A filtering example

```{r}
#| label: filter-example
#| echo: true
#| eval: true

starwars |>
  filter(species == "Human", height > 180) |>
  select(name, height, mass)
```

. . .

::: {.task}
Commas in `filter()` mean AND. Use `|` explicitly when you want OR.
:::

# Conditional values {background-color="#4C326A"}

## From a condition to a value

A comparison gives you a vector of `TRUE` and `FALSE`.

. . .

Often you want to go one step further and assign each row a [different value]{.highlight-peach} depending on the answer.

. . .

That is what `if_else()` and `case_when()` are for.

## `if_else()`

```{r}
#| label: ifelse-signature
#| echo: true
#| eval: false

if_else(condition, true, false, missing = NULL)
```

. . .

::: {.incremental}
- `condition`: a logical vector
- `true`: value returned where the condition is `TRUE`
- `false`: value returned where the condition is `FALSE`
- `missing`: optional value used where the condition is `NA`
:::

## `if_else()` in action

```{r}
#| label: ifelse-mutate
#| echo: true
#| eval: true

starwars |>
  mutate(size = if_else(height > 180, "tall", "short")) |>
  select(name, height, size) |>
  head(4)
```

## `if_else()` is type-strict

The `true` and `false` branches must return the [same type]{.highlight-peach}.

```{r}
#| label: ifelse-strict
#| echo: true
#| eval: true
#| error: true

if_else(c(TRUE, FALSE), "yes", 0)
```

. . .

::: {.task}
This is a feature, not a bug. Base R's `ifelse()` is more forgiving but can silently change types.
:::

## `case_when()`

When you need [more than two outcomes]{.highlight-peach}, nested `if_else()` becomes unreadable.

. . .

`case_when()` handles multiple conditions in a flat, ordered list:

```{r}
#| label: case-when-shape
#| echo: true
#| eval: false

case_when(
  condition_1 ~ value_1,
  condition_2 ~ value_2,
  ...
  .default = fallback
)
```

## Reading `case_when()`

For each row, R checks conditions [from top to bottom]{.highlight-peach} and returns the value next to the first `TRUE`.

```{r}
#| label: case-when-buckets
#| echo: true
#| eval: true

starwars |>
  mutate(
    size = case_when(
      height < 150 ~ "small",
      height < 200 ~ "medium",
      height >= 200 ~ "large"
    )
  ) |>
  select(name, height, size) |>
  head(4)
```

## `.default` and the catch-all

```{r}
#| label: case-when-default
#| echo: true
#| eval: true

x <- c(-1, 5, 12, NA)
case_when(
  x < 0  ~ "negative",
  x < 10 ~ "small",
  x >= 10 ~ "large",
  .default = "unknown"
)
```

. . .

::: {.task}
Anything not matched by an explicit condition (including `NA`) falls into `.default`. Without it, those values become `NA`.
:::

## Order matters

`case_when()` evaluates conditions in order. Put the most specific ones first:

```{r}
#| label: case-when-order
#| echo: true
#| eval: true

age <- c(2, 15, 40, 70)
case_when(
  age < 13 ~ "child",
  age < 20 ~ "teenager",
  age < 65 ~ "adult",
  .default = "senior"
)
```

. . .

If `age < 65` came first, everyone under 65 would end up as `adult`.

# Dates and times {background-color="#4C326A"}

## Why dates are tricky

::: {.incremental}
- Different written formats, `2026-06-04` vs `04/06/2026` vs `June 4, 2026`
- Time zones and daylight saving time
- Leap years (and leap seconds)
- 30 vs 31 days in a month
- Storing dates as text loses arithmetic
:::

. . .

R has dedicated types so you can do real date math.

## R's date types

| Type | What it stores | Created with |
|------|----------------|--------------|
| `Date` | a calendar date | `as_date()`, `ymd()` |
| `POSIXct` | a date-time, seconds since 1970 | `as_datetime()`, `ymd_hms()` |
| `POSIXlt` | a date-time as a list of components | `as.POSIXlt()` |

. . .

In daily use you will mostly meet `Date` and `POSIXct`.

## The reference moment

R stores a `Date` as the [number of days]{.highlight-peach} since **1970-01-01**.

```{r}
#| label: date-internal
#| echo: true
#| eval: true

d <- as_date("2026-06-04")
d
unclass(d)
```

. . .

A `POSIXct` is stored as seconds since the same reference moment, in UTC.

## The `lubridate` package

[`lubridate`]{.highlight-peach} is the tidyverse package for working with dates and times.

. . .

It is loaded with the tidyverse:

```{r}
#| label: load-lubridate
#| echo: true
#| eval: false

library(tidyverse)
```

. . .

Two main families: **parsers** that turn strings into dates, and **accessors** that pull pieces back out.

## Parsing dates from strings

The function name tells you the order of **y**ear, **m**onth, **d**ay:

```{r}
#| label: parsers
#| echo: true
#| eval: true

ymd("2026-06-04")
dmy("04/06/2026")
mdy("June 4, 2026")
```

. . .

Add `_hms`, `_hm`, or `_h` for date-times:

```{r}
#| label: parser-hms
#| echo: true
#| eval: true

ymd_hms("2026-06-04 14:30:00")
```

## `today()` and `now()`

```{r}
#| label: today-now
#| echo: true
#| eval: true

today()
now()
```

. . .

::: {.task}
`today()` returns a `Date`. 

`now()` returns a `POSIXct` with your system's time zone.
:::

## Extracting components

| Function | Returns |
|----------|---------|
| `year()` | the year |
| `month()` | the month (number, or label) |
| `day()` | the day of the month |
| `hour()` | the hour |
| `minute()` | the minute |
| `second()` | the second |

```{r}
#| label: extract
#| echo: true
#| eval: true

d <- ymd("2026-06-04")
year(d)
month(d)
day(d)
```

## `month()` with labels

```{r}
#| label: month-label
#| echo: true
#| eval: true

d <- ymd("2026-06-04")
month(d, label = TRUE)
month(d, label = TRUE, abbr = FALSE)
```

## Day of week, day of year

```{r}
#| label: wday-yday
#| echo: true
#| eval: true

d <- ymd("2026-06-04")

wday(d)
wday(d, label = TRUE)
yday(d)
```

. . .

::: {.task}
`wday()` numbers weeks from Sunday by default. Use `week_start = 1` to start on Monday.
:::

## Converting between types

```{r}
#| label: convert
#| echo: true
#| eval: true

as_date("2026-06-04 14:30:00")
as_datetime("2026-06-04")
```

. . .

`as_date()` drops the time. 

`as_datetime()` lifts a date into a date-time at midnight UTC.

## Time zones with `with_tz()`

`with_tz()` shows the [same instant]{.highlight-peach} in another time zone:

```{r}
#| label: with-tz
#| echo: true
#| eval: true

t <- ymd_hms("2026-06-04 14:30:00", tz = "Europe/Zurich")
t
with_tz(t, "America/New_York")
```

. . .

::: {.task}
Use `force_tz()` if you want to **keep the clock reading** the same and change the zone label, for example to fix data that was tagged with the wrong zone.
:::

## Date arithmetic

Subtracting two dates gives a [duration]{.highlight-peach}:

```{r}
#| label: subtract
#| echo: true
#| eval: true

ymd("2026-07-15") - ymd("2026-06-04")
```

. . .

Add days, weeks, or months with `lubridate` helpers:

```{r}
#| label: add-time
#| echo: true
#| eval: true

ymd("2026-06-04") + days(10)
ymd("2026-06-04") + weeks(2)
ymd("2026-06-04") + months(1)
```

# Displaying tables {background-color="#4C326A"}

## Plain printing is not enough

By default, R prints a data frame as plain text:

```{r}
#| label: plain-print
#| echo: true
#| eval: true

starwars |>
  select(name, height, mass, species) |>
  head(3)
```

. . .

Fine while exploring, not what you want in a [report or paper]{.highlight-peach}.

## Two table packages

:::: {.columns}

::: {.column width="50%"}
**`knitr::kable()`**

- Already there with every Quarto setup
- Quick, minimal, works in PDF and HTML
- Few formatting options
:::

::: {.column width="50%"}
**`gt::gt()`**

- A grammar of tables, similar to ggplot
- Rich formatting: spanners, footnotes, styles
- Best output in HTML
:::

::::

. . .

For everyday reports, reach for `kable()`. For publication-quality tables in HTML, reach for `gt`.

## `knitr::kable()`

```{r}
#| label: kable-basic
#| echo: true
#| eval: true

starwars |>
  select(name, height, mass, species) |>
  head(4) |>
  knitr::kable()
```

## Key `kable()` arguments

| Argument | What it controls |
|----------|------------------|
| `x` | the data |
| `format` | `"html"`, `"latex"`, `"pipe"`, `"markdown"` |
| `digits` | number of decimal places |
| `col.names` | custom column labels |
| `align` | column alignment: `"l"`, `"c"`, `"r"` |
| `caption` | table caption |

## `kable()` with formatting

```{r}
#| label: kable-formatted
#| echo: true
#| eval: true

starwars |>
  select(name, height, mass, species) |>
  head(4) |>
  knitr::kable(
    col.names = c("Name", "Height (cm)", "Mass (kg)", "Species"),
    align     = c("l", "r", "r", "l"),
    digits    = 1
  )
```

## `gt::gt()`

```{r}
#| label: gt-basic
#| echo: true
#| eval: true

starwars |>
  select(name, height, mass, species) |>
  head(4) |>
  gt()
```

## The anatomy of a gt table

A `gt` table has named parts you can address one at a time:

::: {.incremental}
- **table header**: title and subtitle
- **column labels**: the column headers
- **stub**: a row-label column on the left
- **body**: the data cells
- **footer**: footnotes and source notes
:::

## A more polished gt table

```{r}
#| label: gt-polished
#| echo: true
#| eval: true

starwars |>
  select(name, height, mass, species) |>
  head(4) |>
  gt() |>
  tab_header(
    title    = "Star Wars characters",
    subtitle = "First four rows"
  ) |>
  cols_label(
    name    = "Name",
    height  = "Height (cm)",
    mass    = "Mass (kg)",
    species = "Species"
  ) |>
  fmt_number(columns = c(height, mass), decimals = 1)
```

## Useful gt helpers {.smaller}

| Function | Purpose |
|----------|---------|
| `tab_header()` | title and subtitle |
| `cols_label()` | rename columns for display |
| `fmt_number()` | format numbers |
| `fmt_percent()` | format percentages |
| `fmt_currency()` | format currency |
| `tab_spanner()` | grouping label across columns |
| `tab_footnote()` | footnote attached to a cell |
| `tab_source_note()` | source line at the bottom |
| `gtsave()` | save as HTML, PNG, or PDF |

# Captions, labels, cross-references {background-color="#4C326A"}

## Why label a table?

In a long document, you often write something like:

. . .

> "See Table 2 for the breakdown by species."

. . .

Quarto can [number tables automatically]{.highlight-peach} and turn `@tbl-species` into "Table 2", with a live link. You just need to label them properly.

## Chunk options for a table

In any code chunk that produces a table, set two options:

````
```{{r}}
#| label: tbl-species
#| tbl-cap: "Average height by species"

starwars |>
  group_by(species) |>
  summarise(height = mean(height, na.rm = TRUE)) |>
  knitr::kable()
```
````

. . .

::: {.incremental}
- `label` **must** start with `tbl-` for Quarto to treat it as a table
- `tbl-cap` is the caption text
:::

## Cross-referencing in text

In the document body, refer to the table with `@tbl-species`:

. . .

```markdown
Average height differs noticeably between species
(see @tbl-species).
```

. . .

Quarto renders this as:

. . .

> Average height differs noticeably between species (see Table 1).

## Captions for gt tables

`gt` recognises the chunk's `tbl-cap` option, so the same pattern works:

````
```{{r}}
#| label: tbl-starwars-gt
#| tbl-cap: "Star Wars characters"

starwars |>
  select(name, height, mass, species) |>
  head(4) |>
  gt()
```
````

. . .

::: {.task}
Use the chunk options for the caption, not `tab_header()`, when you want Quarto to handle the numbering and the cross-reference.
:::

## Sub-captions for several tables

You can lay out [several tables side by side]{.highlight-peach} with sub-labels:

````
```{{r}}
#| label: tbl-starwars
#| tbl-cap: "Star Wars summaries"
#| tbl-subcap:
#|   - "By species"
#|   - "By homeworld"
#| layout-ncol: 2

starwars |> count(species)   |> knitr::kable()
starwars |> count(homeworld) |> knitr::kable()
```
````

. . .

Cross-reference the pair with `@tbl-starwars`, or a single sub-table with `@tbl-starwars-1`.

## A checklist for nice tables

::: {.incremental}
- Pick `kable()` for quick, lightweight tables
- Pick `gt()` for tables that need careful formatting: titles, footnotes, or custom number formats
- Always set `label` (starting with `tbl-`) and `tbl-cap`
- Cross-reference with `@tbl-...`, never by typing "Table 2" by hand
- Format numbers so the reader can compare them at a glance
:::

## Learn more

::: {.learn-more}
[https://lubridate.tidyverse.org](https://lubridate.tidyverse.org)
:::

::: {.learn-more}
[https://gt.rstudio.com](https://gt.rstudio.com)
:::

::: {.learn-more}
[https://quarto.org/docs/authoring/cross-references.html](https://quarto.org/docs/authoring/cross-references.html)
:::

## {background-color="#4C326A"}

::: {.hand-purple-large}
Questions?
:::
