---
title: "Data transformation"
subtitle: "Data Science Flipped Classroom - Module 3"
author: "Adriana Clavijo Daza"
date: "2026-04-23"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    chalkboard: true
footer: "DSFC - Module 3"
editor: source
execute:
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

# Base R vs packages {background-color="#4C326A"}

## What is base R?

[Base R]{.highlight-peach} is the set of functions and tools that come installed with R itself.

. . .

It includes:

::: {.incremental}
- Core data types: vectors, lists, data frames
- Basic operations: arithmetic, logic, subsetting
- Built-in functions: `mean()`, `sum()`, `length()`, `summary()`
- Built-in datasets: `iris`, `mtcars`, `airquality`
:::

## What is a package?

A [package]{.highlight-yellow} is a collection of functions, data, and documentation that extends R's capabilities.

. . .

::: {.incremental}
- Developed by the R community
- Installed from CRAN, GitHub, or other sources
- Must be loaded before use
:::

## Installing and loading packages

Install a package **once** with `install.packages()`:

```{r}
#| label: install-pkg
#| echo: true
#| eval: false

install.packages("dplyr")
```

. . .

Load it **in every session** with `library()`:

```{r}
#| label: load-pkg
#| echo: true
#| eval: false

library(dplyr)
```

## The tidyverse

The [tidyverse]{.highlight-peach} is a collection of packages designed for data science that share a common philosophy.

. . .

Core packages include: `ggplot2`, `dplyr`, `tidyr`, `readr`, `tibble`, `purrr`, `stringr`, `forcats`, `lubridate`.

. . .

```{r}
#| label: load-tidyverse
#| echo: true
#| eval: false

library(tidyverse)
```

# R building blocks {background-color="#4C326A"}

## Objects

An [object]{.highlight-peach} is anything you store in R's memory with a name.

. . .

```{r}
#| label: objects
#| echo: true
#| eval: true

x <- 10
name <- "Adriana"
numbers <- c(1, 2, 3, 4, 5)
```

. . .

Once created, you can use the name to access the value.

```{r}
#| label: use-object
#| echo: true
#| eval: true

numbers
```

## Functions

A [function]{.highlight-yellow} is a reusable block of code that performs a task.

. . .

```{r}
#| label: function-example
#| echo: true
#| eval: true

mean(numbers)
```

. . .

::: {.task}
Functions are called by their name followed by parentheses: `function_name()`.
:::

## Arguments

[Arguments]{.highlight-peach} are the inputs you pass to a function.

. . .

```{r}
#| label: args-example
#| echo: true
#| eval: true

round(3.14159, digits = 2)
```

. . .

Arguments can be named (`digits = 2`) or positional. Named arguments make your code clearer.

## The assignment operator

Use `<-` to assign a value to an object.

```{r}
#| label: assign
#| echo: true
#| eval: true

age <- 30
age
```

. . .

You can also use `=`, but `<-` is the convention in R.

. . .

::: {.task}
**Keyboard shortcut**

`Alt + -` (Windows/Linux) or `Option + -` (Mac) inserts `<-`.
:::

## The pipe operator

The [pipe]{.highlight-peach} takes the result on the left and passes it as the first argument to the function on the right.

. . .

Without a pipe:

```{r}
#| label: no-pipe
#| echo: true
#| eval: true

round(mean(c(1, 2, 3, 4, 5)), digits = 2)
```

. . .

With a pipe, the same operation reads left to right.

## Two pipe flavors

:::: {.columns}

::: {.column width="50%"}
**magrittr pipe** `%>%`

Comes from the `magrittr` package, loaded with the tidyverse.

```{r}
#| label: magrittr-pipe
#| echo: true
#| eval: true

c(1, 2, 3, 4, 5) %>%
  mean() %>%
  round(digits = 2)
```
:::

::: {.column width="50%"}
**base pipe** `|>`

Built into R since version 4.1 (2021). No package needed.

```{r}
#| label: base-pipe
#| echo: true
#| eval: true

c(1, 2, 3, 4, 5) |>
  mean() |>
  round(digits = 2)
```
:::

::::

## Which pipe to use?

::: {.incremental}
- For new code, prefer the [base pipe]{.highlight-yellow} `|>`: no dependencies, part of R itself
- You will see `%>%` in older code and tutorials: they behave almost the same for our purposes
:::

. . .

::: {.task}
**Keyboard shortcut**

`Ctrl + Shift + M` (Windows/Linux) or `Cmd + Shift + M` (Mac) inserts a pipe. You can set which one in RStudio preferences.
:::

# Introducing dplyr {background-color="#4C326A"}

## What is dplyr?

[dplyr]{.highlight-peach} is the tidyverse package for **data processing**.

. . .

It gives you a consistent set of verbs to manipulate data frames: filter rows, select columns, create new variables, compute summaries, and more.

## The four dplyr rules

::: {.incremental}
1. The **first argument** is always a data frame
2. The **next arguments** describe what to do with that data frame
3. It **returns a data frame**
4. It **does not modify** the original data frame
:::

. . .

::: {.task}
These rules are what make dplyr play beautifully with the pipe.
:::

## Rule 1 and 2: data first, then instructions

```{r}
#| label: rule-1-2
#| echo: true
#| eval: true

filter(starwars, species == "Droid")
```

. . .

With the pipe, the data frame moves out of the call:

```{r}
#| label: rule-1-2-pipe
#| echo: true
#| eval: false

starwars |>
  filter(species == "Droid")
```

## Rule 3: always returns a data frame

This means you can [chain]{.highlight-peach} dplyr functions with the pipe.

```{r}
#| label: rule-3
#| echo: true
#| eval: true

starwars |>
  filter(species == "Human") |>
  select(name, height, mass)
```

## Rule 4: does not modify the object

```{r}
#| label: rule-4
#| echo: true
#| eval: true

starwars |>
  filter(species == "Droid")
```

. . .

The `starwars` object is unchanged. To keep the result, [assign it]{.highlight-yellow} to a new object.

```{r}
#| label: rule-4-assign
#| echo: true
#| eval: true

droids <- starwars |>
  filter(species == "Droid")
```

# The dplyr verbs {background-color="#4C326A"}

## `filter()`: keep rows that match a condition

```{r}
#| label: filter-basic
#| echo: true
#| eval: true

starwars |>
  filter(height > 200)
```

## `filter()` with comparison operators

| Operator | Meaning |
|----------|---------|
| `==` | equal to |
| `!=` | not equal to |
| `<`, `<=` | less than (or equal) |
| `>`, `>=` | greater than (or equal) |
| `%in%` | value is in a set |
| `is.na()` | missing value |

## `filter()`: equality and inequality

```{r}
#| label: filter-eq
#| echo: true
#| eval: true

starwars |>
  filter(eye_color == "blue")
```

## `filter()` with `%in%`

Use `%in%` to match **several values** at once.

```{r}
#| label: filter-in
#| echo: true
#| eval: true

starwars |>
  filter(species %in% c("Droid", "Wookiee", "Ewok"))
```

## `filter()` with multiple conditions

Combine conditions with `&` (and), `|` (or), or a comma (same as `&`).

```{r}
#| label: filter-multi
#| echo: true
#| eval: true

starwars |>
  filter(species == "Human", height > 180)
```

## `select()`: keep or drop columns

```{r}
#| label: select-basic
#| echo: true
#| eval: true

starwars |>
  select(name, height, mass)
```

## `select()` with a minus sign

Use `-` to **drop** columns.

```{r}
#| label: select-minus
#| echo: true
#| eval: true

starwars |>
  select(-films, -vehicles, -starships)
```

## `select()` with a range

Use `:` to select a **range** of consecutive columns.

```{r}
#| label: select-range
#| echo: true
#| eval: true

starwars |>
  select(name:eye_color)
```

## `select()` with helper functions

Functions like `starts_with()`, `ends_with()`, and `contains()` match column names by pattern.

```{r}
#| label: select-starts
#| echo: true
#| eval: true

starwars |>
  select(name, starts_with("h"))
```

. . .

::: {.task}
Other useful helpers: `ends_with()`, `contains()`, `matches()`, `where()`.
:::

## `arrange()`: sort rows

```{r}
#| label: arrange-basic
#| echo: true
#| eval: true

starwars |>
  arrange(height)
```

## `arrange()` in descending order

Wrap a variable in `desc()` to sort from high to low.

```{r}
#| label: arrange-desc
#| echo: true
#| eval: true

starwars |>
  arrange(desc(height))
```

## `relocate()`: change column order

By default, `relocate()` moves columns to the **front**.

```{r}
#| label: relocate-basic
#| echo: true
#| eval: true

starwars |>
  relocate(species, homeworld)
```

## `relocate()` with `.before` and `.after`

```{r}
#| label: relocate-after
#| echo: true
#| eval: true

starwars |>
  relocate(species, .after = name)
```

## `rename()`: change column names

New name goes on the **left**, old name on the **right**.

```{r}
#| label: rename-basic
#| echo: true
#| eval: true

starwars |>
  rename(character = name, planet = homeworld)
```

## `mutate()`: create or modify columns

```{r}
#| label: mutate-basic
#| echo: true
#| eval: true

starwars |>
  mutate(height_m = height / 100) |>
  select(name, height, height_m)
```

## `mutate()`: multiple columns at once

```{r}
#| label: mutate-multi
#| echo: true
#| eval: true

starwars |>
  mutate(
    height_m = height / 100,
    bmi = mass / (height_m ^ 2)
  ) |>
  select(name, height_m, mass, bmi)
```

## `transmute()`: mutate and keep only new columns

```{r}
#| label: transmute-basic
#| echo: true
#| eval: true

starwars |>
  transmute(
    name,
    height_m = height / 100,
    bmi = mass / (height / 100) ^ 2
  )
```

::: {.task}
`transmute()` is like `mutate()` followed by `select()`, all in one step.
:::

## `summarize()`: collapse to summary statistics

```{r}
#| label: summarize-basic
#| echo: true
#| eval: true

starwars |>
  summarize(
    avg_height = mean(height, na.rm = TRUE),
    avg_mass = mean(mass, na.rm = TRUE),
    n = n()
  )
```

::: {.task}
`summarize()` returns one row per group, with the columns you define.
:::

## `group_by()`: compute by group

`group_by()` on its own does not change the data, but it tells dplyr that subsequent operations should run **per group**.

```{r}
#| label: group-by
#| echo: true
#| eval: true

starwars |>
  group_by(species) |>
  summarize(
    n = n(),
    avg_height = mean(height, na.rm = TRUE)
  )
```

## `group_by()` with multiple variables

```{r}
#| label: group-by-multi
#| echo: true
#| eval: true

starwars |>
  group_by(species, gender) |>
  summarize(n = n(), .groups = "drop")
```

## `count()`: shortcut for grouped counts

`count(x)` is equivalent to `group_by(x) |> summarize(n = n())`.

```{r}
#| label: count-basic
#| echo: true
#| eval: true

starwars |>
  count(species)
```

## `count()` with sorting

```{r}
#| label: count-sort
#| echo: true
#| eval: true

starwars |>
  count(species, sort = TRUE)
```

# The big picture {background-color="#4C326A"}

## Verbs at a glance {.smaller}

| Verb | What it does |
|------|--------------|
| `filter()` | Keep rows that match a condition |
| `select()` | Keep or drop columns |
| `arrange()` | Sort rows |
| `relocate()` | Reorder columns |
| `rename()` | Rename columns |
| `mutate()` | Create or modify columns |
| `transmute()` | Create columns, drop the rest |
| `summarize()` | Collapse to summary statistics |
| `group_by()` | Compute per group |
| `count()` | Shortcut for grouped counts |

## Putting it all together

```{r}
#| label: pipeline
#| echo: true
#| eval: true
#| code-line-numbers: "|1|2|3|4|5|6"

starwars |>
  filter(!is.na(height), !is.na(mass)) |>
  mutate(bmi = mass / (height / 100) ^ 2) |>
  group_by(species) |>
  summarize(avg_bmi = mean(bmi), n = n()) |>
  arrange(desc(avg_bmi))
```

## {background-color="#4C326A"}

::: {.hand-purple-large}
Questions?
:::
