---
title: "Module 5 exercise"
subtitle: "Conditions, dates, tables, figures"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
  eval: false
---

# Atlantic hurricanes

In this exercise you will pull together everything you have learned so
far in the course (reading data, transforming, conditions, dates, tables
and figures) on the `storms` dataset that comes with the `dplyr`
package.

The dataset records the position and characteristics of Atlantic
storms between 1975 and 2022. A copy is provided at
`../data/storms.csv` so you can practise reading it from disk.

Fill in each code chunk where you see `___`. Render the document when
you are done and check that the table and the figure are numbered, and
that the cross-references in the last section turn into clickable links.

## Setup

Run the chunk below first. It loads every package you will need.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(gt)
library(knitr)
```

## 1. Read the data

The data file is at `../data/storms.csv`. It is a regular comma-separated
file, no tricks. Read it with `read_csv()` and store the result in an
object called `storms`.

```{r}
#| label: read-data

storms <- read_csv(___)

# A quick check of the columns and their types
glimpse(storms)
```

::: {.callout-tip}
Use `glimpse()` to get an overview of the data.
:::

## 2. Filter

For the rest of the exercise you will work with the rows where the
storm reached **hurricane** status, between **2010 and 2022**
(inclusive).

Fill in the `filter()` call below so that:

- `status` equals `"hurricane"`
- `year` is at least 2010 and at most 2022

```{r}
#| label: filter

hurricanes <- storms |>
  filter(___)

# How many rows did you end up with?
nrow(hurricanes)
```

## 3. Mutate: a date and an intensity label

You will now add three new columns to `hurricanes`:

- `date`: a proper `Date` built from `year`, `month` and `day`. Use the
  `lubridate` helper `make_date()`.
- `month_name`: the month as a labelled factor. Use `month()` with
  `label = TRUE` and `abbr = FALSE` for the full name.
- `intensity`: a label that classifies each row by `wind` (in knots),
  following the Saffir-Simpson scale:

  | Condition       | Label    |
  |-----------------|----------|
  | `wind < 83`     | `"Cat 1"`|
  | `wind < 96`     | `"Cat 2"`|
  | `wind < 113`    | `"Cat 3"`|
  | `wind < 137`    | `"Cat 4"`|
  | otherwise       | `"Cat 5"`|

  Use `case_when()` to build the labels, then turn the result into a
  **factor** with `factor(..., levels = ...)`. Setting `levels` in the
  right order is what keeps the categories sorted from weakest to
  strongest in tables and plots: no need to make it `ordered = TRUE`
  unless you actually want to compare categories with `<` or `>`.

```{r}
#| label: mutate

hurricanes <- hurricanes |>
  mutate(
    date       = ___,
    month_name = ___,
    intensity  = case_when(
      ___
    ),
    intensity  = factor(intensity, levels = ___)
  )

hurricanes |>
  select(name, date, month_name, wind, intensity) |>
  head(6)
```

::: {.callout-tip}
After you finish this step, check that `date` is of class `Date`, and
that `intensity` is a factor whose `levels()` are the five
Saffir-Simpson categories in the right order.
:::

## 4. A summary table

Group `hurricanes` by `intensity` and compute, for each category:

- the number of observations, using `n()`
- the mean wind speed, using `mean(wind, na.rm = TRUE)`
- the minimum atmospheric pressure, using `min(pressure, na.rm = TRUE)`

Then **sort the result by mean wind speed in descending order** with
`arrange(desc(mean_wind))`, so the strongest categories appear at the
top.

Display the result with either `knitr::kable()` or `gt::gt()`. The chunk
options below already set a `tbl-` label and a `tbl-cap`, so the table
will be numbered and you can refer to it later as `@tbl-intensity`.

```{r}
#| label: tbl-intensity
#| tbl-cap: "Atlantic hurricanes 2010-2022 by Saffir-Simpson category, sorted by mean wind speed"

hurricanes |>
  group_by(___) |>
  summarise(
    n            = ___,
    mean_wind    = ___,
    min_pressure = ___
  ) |>
  arrange(___) |>
  # here add code to print table
```

::: {.callout-tip}
Use `knitr::kable(digits = 1)` or `gt()` with `tab_header()`, `cols_label()`, and `fmt_number()`. Keep the chunk label and caption as they are.
:::

## 5. A visualization

Plot the **storm tracks** on a map of the Atlantic basin.

Build the plot step by step:

- Country outlines come from `map_data("world")`, which needs the
  `maps` package (install with `install.packages("maps")` if you do
  not have it yet). The first line of the chunk loads them into a
  data frame called `world`.
- In the `ggplot()` call, map `long` to `x` and `lat` to `y`.
- Draw the coastline first with `geom_polygon()`, using the `world`
  data, so the storm tracks are drawn on top of it.
- Use `geom_path()` to connect the positions of each storm, with
  `aes(group = name)` so each storm gets its own line. Color the
  lines by `intensity`.
- `coord_quickmap()` already has reasonable `xlim` and `ylim` for the
  Atlantic basin: leave those in place.
- Give the plot a title and informative axis labels.

The chunk options already set a `fig-` label and a `fig-cap`, so you
will be able to refer to the figure as `@fig-tracks`.

```{r}
#| label: fig-tracks
#| fig-cap: "Tracks of Atlantic hurricanes between 2010 and 2022, overlaid on a coastline outline and coloured by Saffir-Simpson category at each point along the track."
#| fig-width: 8
#| fig-height: 5

world <- map_data("world")

ggplot(hurricanes, aes(x = ___, y = ___)) +
  geom_polygon(
    data = world,
    aes(x = long, y = lat, group = group),
    fill = "grey92", color = "white"
  ) +
  geom_path(aes(group = ___, color = ___), alpha = 0.6) +
  coord_quickmap(xlim = c(-110, -10), ylim = c(5, 55)) +
  labs(
    title = ___,
    x     = "Longitude",
    y     = "Latitude",
    color = "Category"
  ) +
  theme_minimal()
```

## 6. Write your findings

Replace the two short paragraphs below with your own observations on
the table and the figure. **Keep the `@tbl-...` and `@fig-...`
references**, that is what makes Quarto turn them into "Table 1" and
"Figure 1" with clickable links.

- Loking at @tbl-... 
- @fig-... shows...
